The Name of the Title Is Hope

### 2024-12-30 paper.tex
still churnin'

\section{Background}
\subsection{The GPU Programming Environment(Model?)}
\john{So I have written, roughly, a zillion of these sections and here is my current philosophy. The danger is that you write something really interesting and relevant and a reviewer who is familiar with GPUs sees the title and sees the first few lines and says ``I can skip this'' and does. There are plenty of great descriptions of how GPUs work. Cite one of them. What this section needs to do is very specifically call out the interesting/detailed features of GPUs that are \emph{particularly relevant to this paper} that are not just ``here's generally how a GPU works''. For me, that's the last paragraph. The other part that might be necessary is defining the specific terminology used in the paper (e.g., workgroup vs.\ thread block vs.\ CTA). If you feel like breaking up a wall of text, which I generally think is a good idea, we can make a little table of equivalent terminology and boldface the terms we're using.} \john{Perhaps another good thing to mention that is specific to this paper is the GPU's support for synchronization. I could see having a blurb on ``there is no global inter-workgroup synchronization except an expensive kernel barrier'' and also what \emph{is} available in terms of pairwise synchronization between workgroups and what the requirements are (e.g., residence). These are more detailed (and important) GPU topics that aren't present in the higher-level summary you have below. Later in the paper, you reference ``without an inter-workgroup barrier primitive'' and don't say why, and you don't want the reader to get the wrong idea here (you don't want the reader to say ``oh, we can just add this to the language, we're done'', and you alternatively don't want the reader to say ``oh, we'll just use a kernel launch barrier''). There is a fundamental reason why this primitive is expensive. Here's where you can discuss that, and then assert that a global barrier (presumably a kernel launch barrier) is so expensive that we do not consider it in our implementations.} \john{Thinking about the above, one way to organize this subsection that is most helpful is to \emph{show} here what kind of synchronizations are / aren't possible within the GPU programming model, in a principled way, and then these are the constraints for our implementation. We want readers to not just accept a weird set of constraints as a given but rather to understand where they come from, and why.}
Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency. Due to their specialized nature, GPUs function as co-processors to the CPU, requiring GPU programs—known as \emph{kernels} or \emph{shaders}—to be \emph{launched} or \emph{dispatched} heterogeneously by the CPU\@. Upon launch, a kernel is replicated across independent execution contexts, called threads. These threads are first organized into smaller units known as \emph{subgroups}\footnote{Also referred to as \emph{Warps}, \emph{Waves}, or \emph{SIMD-Groups}.}, which are then grouped into larger units called \emph{workgroups}\footnote{Also referred to as \emph{Cross-Thread-Arrays}, \emph{Thread-Blocks}, \emph{Blocks}, \emph{Thread-Groups}, or \emph{Groups}.}. This thread hierarchy mirrors the GPU memory hierarchy: individual threads access private, high-speed registers; threads within a workgroup share low-latency shared memory\footnote{Also referred to as \emph{Local Data Storage}, \emph{Group-Shared Memory}, (Check metal terminology).}; and all threads have access to high-bandwidth but higher-latency global memory.

@@

\subsection{The Scan Primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{10.1145/322217.322232, 5219801}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid~\cite{}. \john{Rabin-Karp's second fingerprinting method for string matching (``Efficient randomized pattern-matching algorithms'', Section 6, 1987)~\cite[Section 6]{Karp:1987:ERP} has 2$\times$2 matrix multiplication as its binary operation, which is non-commutative, and is a good thing to cite in a paper going to a theory conference like SPAA, because string matching is a super-SPAA-friendly topic. I put it in the bib if you want to use it.} In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence. If the subset includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]
Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris offers a taxonomy that relates depth, fanout, and wire tracks~\cite{}, while Hinze develops an algebraic framework for scans~\cite{}. Snir proved that \emph{depth} and \emph{size} are related by (fill in)~\cite{}, and Fich proved that among minimum depth scans Ladner-Fischer networks have optimal size. Finally, Merrill and Garland provide a comprehensive review of GPU scan implementations. In this work, we focus on the evolution of inter-workgroup (global) scan strategies and evaluate their suitability for our ideal architecture.

@@

\john{So if I was explaining this evolution without having read this paper, the way I would do it is by asserting that scan has a relatively low arithmetic intensity and we expect theoretically, and see experimentally, that scans are entirely limited by memory bandwidth, and in fact can run near the theoretical speed of light of the memory system. Given that, the entire goal is to make sure we organize computation in a way that allows us to use all of the theoretical memory bandwidth. Given THAT, we can talk about using the memory and compute hierarchy (shared memory, tiles, etc.) to make sure we do.}

@@

@@

\subsubsection{Chained-Scan}
Continuing the trend of global data movement optimization, the most recent work on global scan has been the \emph{Chained-Scan} architecture. First introduced by Yan et al.\ in \emph{StreamScan}, the key innovation of GPU chained scanning lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups. This shifts the problem from explicit synchronization to implicit synchronization via the order in which workgroups access global data. Each workgroup processes its assigned tile independently, waiting only for the result of the preceding reductions before continuing. This structure ensures that global memory traffic is minimized, as each element is read and written exactly once, achieving the theoretical minimum of $2n$.

While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving speed-of-light performance. This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the reduction of the preceding workgroup. Although this reduces global memory access during the propagation phase to $O(n/t)$, the sequential dependency creates a bottleneck, constraining the algorithm’s performance to the rate at which reductions propagate through device memory. Despite this limitation, the total data movement remains optimal at $O(2n + n/t) = O(2n)$.

@@

In addition to its speed, the \emph{Chain-Scan} architecture has other, more practical properties, that improve over its predecessors. As it is a single kernel, it minimizes launch overheads and simplifies maintenance. On previous architectures, handling of arbitrary inputs can be awkward: the recursive approach, for example, entails multiple kernel launches and a heterogenous system to coordinate them. Similarly, a raking approach can be cumbersome, as determining the workgroup occupancy for every target device is often impractical. Furthermore, on GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—raking the workgroups can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. As a consequence of its $2n$ global data movement and inherently serial ordering, \emph{Chained-Scan} is also suitable for \emph{in-situ} compaction and allocation. In previous architectures, such tasks often required additional kernel launches or complex bookkeeping to handle dependencies between workgroups. By contrast, \emph{Chained-Scan} efficiently propagates dependencies within a single kernel, minimizing overhead and enabling seamless integration of compaction and allocation operations.

Thus, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture, it falls short in meeting the criterion of \emph{portability}. As we discuss in more detail in the following section, this limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA hardware but not universally available across all vendors and architectures. Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure, making it unsuitable for the inherently cross-vendor, cross-architecture environment of shading languages.

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
A \emph{forward-progress guarantee} (FPG) is a promise provided by the subgroup scheduler that all occupant subgroups eventually make progress toward termination, preventing scenarios where a subgroup is starved of execution time. Sorensen et al.~\cite{} and Sorensen et al.~\cite{} provide the most comprehensive analysis of GPU schedulers, including the implications of running FPG dependent algorithms without it.

@@

For instance, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor. Although the multiprocessor can host both workgroups, it has only one scheduling unit, which lacks FPG\@. In this setup, both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins waiting for the reduction of \emph{Work Tile 0}. However, because the scheduler lacks FPG, it may continue scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---it remains absent on some contemporary hardware, most notably certain ARM-based chips, including Apple's M-series GPUs~\cite{10.1145/3485508}. Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad\footnote{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time. It's also unclear why the test did not trigger Metal's timeout device recovery}, and at worst, risks the afformentioned TDR and subsequent program crash. This extreme slowdown highlights the inherent fragility of \emph{Chained-Scan} without FPG\@.

Exacerbating this issue is the fact that no major graphics API currently provides developers with a mechanism to query FPG support\footnote{As of writing this, 12/29/2024, we are unaware of any major graphics API---Vulkan, D3D12, or Metal---that provide developers with an FPG capability query}. This lack of transparency forces developers to design algorithms without clear knowledge of whether critical guarantees are available on the target hardware. In environments like shading languages, where cross-platform portability is required, this uncertainty compels developers to revert to older, less efficient strategies such as \emph{Reduce-Then-Scan}. However, as discussed earlier, \emph{Reduce-Then-Scan} carries its own set of drawbacks, including increased global memory traffic ($O(3n)$) and additional kernel launches, which result in reduced throughput and higher latency. While these tradeoffs were once deemed acceptable, they stand in stark contrast to the efficiency and simplicity of \emph{Chained-Scan} when implemented on hardware with reliable FPG\@. Yet, in the absence of guaranteed forward progress, developers have little choice but to sacrifice performance for broader compatibility.

\subsection{Earlier Attempts at Portability (Cuttable?)}
Levien and Naur~\cite{}, made the first attempt at solving forward progress in \emph{DecoupledLookback}, introducing \emph{ScallarFallback}. In \emph{ScallarFallback}, each time a workgroup waits on a preceding tile, it begins a \emph{fallback} operation, and begins consuming the blocking tile one element at a time. Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because a workgroup can also perform the reduction of a blocking tile itself, thus breaking the inter-workgroup dependency.

\emph{ScallarFallback} suffers from two primary issues that impact its efficiency. First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup engages in a waiting spin.This effectively takes $t$ spins to complete the fallback of the deadlocking tile. Moreover, because only a single thread is responsible for loading elements, the cache line is severely underutilized, further degrading performance. Second, once the fallback is complete, the workgroup does not post its result to device memory. Consequently, every workgroup deadlocked by that tile must redundantly execute the fallback, wasting computational resources. While \emph{ScallarFallback} was able to achieve speed of light on hardware supporting FPG and executed correctly on hardware without FPG, its performance on the latter was inferior to that of \emph{Reduce-Then-Scan}.

@@

\subsection{Non Goals}
Although our goal is to create a fully portable and highly performant scan implementation, we are constrained by underlying hardwares and programming models. As discussed in more detail in \emph{Limits on the Speed-of-Light}, not all architectures offer atomic operations that are sufficiently fast enough or scheduling models that are sufficeintly fair enough to attain speed-of-light performance, and thus we cannot guarantee such performance. Although we are cognisant of the risks posed by subgroup divergence, subgroup functions in shading languages do not allow explicit divergence control\footnote{For example, CUDA allows developers to explicitly provision subgroup functions with a mask of the threads that will participate in it.}, placing a solution to subgroup divergence issues outside our scope. Furthermore, although real-world scans may require more than 30 bits of space for their reductions, or may involve scanning across multiple struct members, WGSL’s current limitations---lack of fences and 64-bit atomics---lead us to focus on the simplest scenario that fits within the 30-bit range. Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained-Scan}.

>>>>

\section{Background}
\subsection{The GPU Programming Environment(Model?)}
\john{So I have written, roughly, a zillion of these sections and here is my current philosophy. The danger is that you write something really interesting and relevant and a reviewer who is familiar with GPUs sees the title and sees the first few lines and says ``I can skip this'' and does. There are plenty of great descriptions of how GPUs work. Cite one of them. What this section needs to do is very specifically call out the interesting/detailed features of GPUs that are \emph{particularly relevant to this paper} that are not just ``here's generally how a GPU works''. For me, that's the last paragraph. The other part that might be necessary is defining the specific terminology used in the paper (e.g., workgroup vs.\ thread block vs.\ CTA). If you feel like breaking up a wall of text, which I generally think is a good idea, we can make a little table of equivalent terminology and boldface the terms we're using.} \john{Perhaps another good thing to mention that is specific to this paper is the GPU's support for synchronization. I could see having a blurb on ``there is no global inter-workgroup synchronization except an expensive kernel barrier'' and also what \emph{is} available in terms of pairwise synchronization between workgroups and what the requirements are (e.g., residence). These are more detailed (and important) GPU topics that aren't present in the higher-level summary you have below. Later in the paper, you reference ``without an inter-workgroup barrier primitive'' and don't say why, and you don't want the reader to get the wrong idea here (you don't want the reader to say ``oh, we can just add this to the language, we're done'', and you alternatively don't want the reader to say ``oh, we'll just use a kernel launch barrier''). There is a fundamental reason why this primitive is expensive. Here's where you can discuss that, and then assert that a global barrier (presumably a kernel launch barrier) is so expensive that we do not consider it in our implementations.} \john{Thinking about the above, one way to organize this subsection that is most helpful is to \emph{show} here what kind of synchronizations are / aren't possible within the GPU programming model, in a principled way, and then these are the constraints for our implementation. We want readers to not just accept a weird set of constraints as a given but rather to understand where they come from, and why.}
Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency. Due to their specialized nature, GPUs function as co-processors to the CPU, requiring GPU programs—known as \emph{kernels} or \emph{shaders}—to be \emph{launched} or \emph{dispatched} heterogeneously by the CPU\@. Upon launch, a kernel is replicated across independent execution contexts, called threads. These threads are first organized into smaller units known as \emph{subgroups},\footnote{Also referred to as \emph{Warps}, \emph{Waves}, or \emph{SIMD-Groups}.} which are then grouped into larger units called \emph{workgroups}.\footnote{Also referred to as \emph{Cross-Thread-Arrays}, \emph{Thread-Blocks}, \emph{Blocks}, \emph{Thread-Groups}, or \emph{Groups}.} This thread hierarchy mirrors the GPU memory hierarchy: individual threads access private, high-speed registers; threads within a workgroup share low-latency shared memory;\footnote{Also referred to as \emph{Local Data Storage}, \emph{Group-Shared Memory}, (Check metal terminology).} and all threads have access to high-bandwidth but higher-latency global memory.

@@

\subsection{The Scan Primitive}\label{sec:the-scan-primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{10.1145/322217.322232, 5219801}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid~\cite{}. \john{Rabin-Karp's second fingerprinting method for string matching (``Efficient randomized pattern-matching algorithms'', Section 6, 1987)~\cite[Section 6]{Karp:1987:ERP} has 2$\times$2 matrix multiplication as its binary operation, which is non-commutative, and is a good thing to cite in a paper going to a theory conference like SPAA, because string matching is a super-SPAA-friendly topic. I put it in the bib if you want to use it.} In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence. If the subset includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]
Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris offers a taxonomy that relates depth, fanout, and wire tracks~\cite{}, while Hinze develops an algebraic framework for scans~\cite{}. Snir proved that \emph{depth} and \emph{size} are related by (fill in)~\cite{}, and Fich proved that among minimum depth scans Ladner-Fischer networks have optimal size. Finally, Merrill and Garland provide a comprehensive review of GPU scan implementations. In this work, we focus on the evolution of inter-workgroup (global) scan strategies and evaluate their suitability for our ideal architecture.

@@

\john{So if I was explaining this evolution without having read this paper, the way I would do it is by asserting that scan has a relatively low arithmetic intensity and we expect theoretically, and see experimentally, that scans are entirely limited by memory bandwidth, and in fact can run near the theoretical speed of light of the memory system. So the first goal is to find an algorithmic formulation that requires a minimum memory volume. Then given the above, and given a particular algorithmic formulation, the entire goal is to make sure we organize computation in a way that allows us to use all of the theoretical memory bandwidth. Given THAT, we can talk about using the memory and compute hierarchy (shared memory, tiles, etc.) to make sure we do.}

@@

\john{I liked what you said above (Section~\ref{sec:the-scan-primitive}) about how each of these formulations ``navigate this inter-workgroup dependency''. Make that explicit in this subsubsection, and each of the below!}

@@

\subsubsection{Chained-Scan}
Continuing the trend of global data movement optimization, the most recent work on global scan has been the \emph{Chained-Scan} architecture. First introduced by Yan et al.\ in \emph{StreamScan}, the key innovation of GPU chained scanning lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups. This shifts the problem from explicit synchronization to implicit synchronization via the order in which workgroups access global data. Each workgroup processes \john{I think you want to say, explicitly, the work that each workgroup actually does, and note what part of that is latency-sensitive (since, as you say below, the latency of the sequential dependency is super important for overall scan performance.} its assigned tile independently, waiting only for the result of the preceding reductions before continuing. This structure ensures that global memory traffic is minimized, as each element is read and written exactly once, achieving the theoretical minimum of $2n$.

While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving speed-of-light performance. This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the reduction of the preceding workgroup. Although this reduces global memory access during the propagation phase to $O(n/t)$, the sequential dependency creates a bottleneck, constraining the algorithm’s performance to the rate at which reductions propagate through device memory. Despite this limitation, the total data movement remains optimal at $O(2n + n/t) = O(2n)$. \john{Somewhere in here you want to put in at least a hint of ``forward progress guarantee''. You don't have to use those words. But when you get to FPG, you want to point back and say ``this formulation depends on \ldots''.}

@@

In addition to its speed, the \emph{Chain-Scan} \john{Be consistent, Chained-Scan or Chain-Scan?} architecture has other, more practical properties, that improve over its predecessors. As it is a single kernel, it minimizes launch overheads and simplifies maintenance. On previous architectures, handling of arbitrary inputs \john{input sizes?} can be awkward: the recursive approach, for example, entails multiple kernel launches and a heterogenous system to coordinate them. Similarly, a raking approach \john{(a necessary optimization in approaches prior to Chained-Scan) ?} can be cumbersome, as determining the workgroup occupancy for every target device is often impractical. Furthermore, on GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—raking the workgroups can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. As a consequence of its $2n$ global data movement and inherently serial ordering, \emph{Chained-Scan} is also suitable for \emph{in-situ} compaction and allocation. In previous architectures, such tasks often required additional kernel launches or complex bookkeeping to handle dependencies between workgroups. By contrast, \emph{Chained-Scan} efficiently propagates dependencies within a single kernel, minimizing overhead \john{Your audience might not have any idea what you are talking about in the next phrase. Add something like ``augmenting scan with useful extensions such as \ldots''?} and enabling seamless integration of compaction and allocation operations.

But, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture \john{``in terms of its theoretical and implementation characteristics'' ?}, it falls short in meeting the criterion of \emph{portability}. As we discuss in more detail in the following section, this limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA hardware but not universally available across all vendors and architectures. \john{I would probably try to be precise where you say ``not universally available''. That phrase is not to our advantage because it implies that mostly its available and maybe there's this one weirdo GPU where it's not. I want to be accurate, but I also want that phrase to imply (if it's true) that there's a significant volume of devices that lack FPG\@.} Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure, making it unsuitable for the inherently cross-vendor, cross-architecture environment of shading languages.

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
A \emph{forward-progress guarantee} (FPG) is a promise provided by the subgroup scheduler that all occupant subgroups eventually make progress toward termination, preventing scenarios where a subgroup is starved of execution time. Sorensen et al.~\cite{} and Sorensen et al.~\cite{} provide the most comprehensive analysis of GPU schedulers, including the implications of running FPG dependent algorithms without it. \john{It would benefit this paper if you contrasted their work vs.\ the work that is here. For instance, if they don't discuss chained scan / if they focus on a different class of algorithms, that's worth saying. You don't want a reviewer coming back and saying ``oh, they already totally covered the topic of FPG''\@.}

@@

For instance, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor. Although the multiprocessor can host both workgroups, it has only one scheduling unit, which lacks FPG\@. In this setup, both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins waiting \john{I wonder if ``waiting'' needs to be formalized a little here. The casual reader might think ``if Work Tile 1 is `waiting', why is it being scheduled at all?''. I'm presuming it's spinning and thus actually getting scheduled as a result. We want to avoid a reviewer making his/her own conclusions about how that hardware / the implementation SHOULD work and then disagreeing with our conclusions. We need to make it clear this isn't an easily avoidable problem.} for the reduction of \emph{Work Tile 0}. However, because the scheduler lacks FPG, it may continue scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. \john{We also want to avoid a reader thinking ``well, why not just schedule one work tile per multiprocessor, doesn't that solve the problem?''.} As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---it remains absent on some contemporary hardware, most notably certain ARM-based chips, including Apple's M-series GPUs~\cite{10.1145/3485508}. Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad,\footnote{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time. It's also unclear why the test did not trigger Metal's timeout device recovery} and at worst, risks the aforementioned TDR and subsequent program crash. This extreme slowdown highlights the inherent fragility of \emph{Chained-Scan} without FPG\@.

Exacerbating this issue is the fact that no major graphics API currently provides developers with a mechanism to query FPG support.\footnote{As of writing this, 12/29/2024, we are unaware of any major graphics API---Vulkan, D3D12, or Metal---that provide developers with an FPG capability query} This lack of transparency forces developers to design algorithms without clear knowledge of whether critical guarantees are available on the target hardware. In environments like shading languages, where cross-platform portability is required, this uncertainty compels developers to revert to older, less efficient strategies such as \emph{Reduce-Then-Scan}. However, as discussed earlier, \emph{Reduce-Then-Scan} carries its own set of drawbacks, including increased global memory traffic ($O(3n)$) and additional kernel launches, which result in reduced throughput and higher latency. While these tradeoffs were once deemed acceptable, they stand in stark contrast to the efficiency and simplicity of \emph{Chained-Scan} when implemented on hardware with reliable FPG\@. Yet, in the absence of guaranteed forward progress, developers have little choice but to sacrifice performance for broader compatibility.

\subsection{Earlier Attempts at Portability (Cuttable?)}
Levien and Naur~\cite{}, made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress in \emph{DecoupledLookback}, introducing \emph{ScallarFallback}. \john{Scalar, not Scallar?} In \emph{ScallarFallback}, each time a workgroup waits on a preceding tile, it begins a \emph{fallback} operation, and begins consuming the blocking tile one element at a time. Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because a workgroup can also perform the reduction of a blocking tile itself, thus breaking the inter-workgroup dependency.

\emph{ScallarFallback} suffers from two primary issues that impact its efficiency. First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup engages in a waiting spin. This effectively takes $t$ spins to complete the fallback of a deadlocking $t$-sized tile. Moreover, because only a single thread is responsible for loading elements, the cache line associated with the loaded elements is severely underutilized, further degrading performance. Second, once the fallback is complete, the workgroup does not post its result to device memory. Consequently, every workgroup deadlocked by that tile must redundantly execute the fallback, wasting computational resources. While \emph{ScallarFallback} was able to achieve speed of light on hardware supporting FPG and executed correctly on hardware without FPG, its performance on the latter was inferior to that of \emph{Reduce-Then-Scan}.

@@

\subsection{Non Goals}
Although our goal is to create a fully portable and highly performant scan implementation, we are constrained by underlying hardwares and programming models. As discussed in more detail in \emph{Limits on the Speed-of-Light}, not all architectures offer atomic operations that are sufficiently fast enough or scheduling models that are sufficeintly fair enough to attain speed-of-light performance, and thus we cannot guarantee such performance. Although we are cognisant of the risks posed by subgroup divergence, subgroup functions in shading languages do not allow explicit divergence control,\footnote{For example, CUDA allows developers to explicitly provision subgroup functions with a mask of the threads that will participate in it.} placing a solution to subgroup divergence issues outside our scope. Furthermore, although real-world scans may require more than 30 bits of space for their reductions, or may involve scanning across multiple struct members, WGSL’s current limitations---lack of fences and 64-bit atomics---lead us to focus on the simplest scenario that fits within the 30-bit range. Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained-Scan}.
